---
title: "CQRM: Composite Quantile Regression Models for ncRNA Differential Expression"
subtitle: "Simulation Tutorial — Data Generation, Method Comparison, and Results"
author: "Zhangchi Xu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

---

# Overview

This tutorial walks through a complete simulation study using the **CQRM**
package. We will:

1. **Install / load** the CQRM package and dependencies.
2. **Generate** a synthetic kidney-inspired dataset.
3. **Apply** CQRM and comparison methods.
4. **Display** adjusted p-value tables and FDP/power summaries.

All CQRM methods precompute `solve(subsigma)` *outside* the B-loop so that
each weight-pattern inverse is computed only once per null simulation, not once
per null draw.

---

# Setup

## Install CQRM

```{r install, eval=FALSE}
# From GitHub (once the package is public):
remotes::install_github("xzc0907/CQRM")

# Or load from local source:
source("../R/internals.R")
source("../R/CQRM.R")
```

## Load dependencies

```{r load-deps}
# CQRM
if (requireNamespace("CQRM", quietly = TRUE)) {
  library(CQRM)
} else {
  source("../R/internals.R")
  source("../R/CQRM.R")
}

# Comparison methods (loaded from the wrapper script)
source("comparison_methods.R")

# Display helpers
library(knitr)
library(kableExtra)
```

---

# Data Generation

## Synthetic kidney-inspired dataset

The synthetic data mirrors the design of the paper's non-parametric simulation
(`synthetic_nonpara_server_1.R`). It uses the `kidney` RNA-seq dataset from the
**SimSeq** package as a template:

| Parameter | Value | Description |
|-----------|-------|-------------|
| `n_gene`  | 500   | Total number of genes in the synthetic dataset |
| `DE_prop` | 0.10  | Proportion of true differentially expressed (DE) genes |
| `size`    | 50    | Samples per group drawn from kidney (kidney has 72/group) |
| `seed`    | 42    | Random seed for reproducibility |
| `alpha_consensus` | 0.001 | BH threshold for consensus DE gene pool (DESeq2 ∩ Wilcoxon) |

The key steps are:

1. Load and filter genes with >80% zero counts.
2. Run DESeq2 and Wilcoxon on the full kidney data; intersect their discoveries
   at a stringent threshold to define the **true DE pool**.
3. Sample `n_DE = n_gene × DE_prop` genes from the true DE pool.
4. Sample `n_non_DE = n_gene × (1 − DE_prop)` genes from the non-DE pool and
   **permute** their group labels so they are truly null.
5. Stack DE and (permuted) non-DE genes, save as `data/synthetic_kidney.rds`.

```{r generate-data, eval=FALSE}
# Run once to generate and save the dataset (takes ~5 min for full kidney run).
# Skip if data/synthetic_kidney.rds already exists.
source("generate_data.R")
```

## Load generated data

```{r load-data}
dat        <- readRDS("data/synthetic_kidney.rds")
counts     <- dat$counts         # G × (2 * size) integer count matrix
group      <- dat$group          # group vector: 0 = group 1, 1 = group 2
is_true_DE <- dat$is_true_DE     # logical: TRUE for true DE genes
size       <- dat$size           # samples per group

G    <- nrow(counts)
n_DE <- sum(is_true_DE)
cat(sprintf("Genes: %d  (DE: %d, non-DE: %d)  |  Samples/group: %d\n",
            G, n_DE, G - n_DE, size))

count1 <- counts[, group == 0]   # gene × sample matrix, group 1
count2 <- counts[, group == 1]   # gene × sample matrix, group 2
```

## Quick look at the data

```{r data-summary, fig.width=7, fig.height=4}
# Distribution of total counts per sample
sample_totals <- colSums(counts)
hist(sample_totals, breaks = 20, main = "Library sizes (total counts/sample)",
     xlab = "Total counts", col = "steelblue", border = "white")

# Proportion of zeros per gene in group 1
zero_frac <- rowMeans(count1 == 0)
hist(zero_frac, breaks = 20, main = "Zero fraction per gene (group 1)",
     xlab = "Proportion of zero counts", col = "salmon", border = "white")
```

---

# Running the Methods

## Parameters

```{r params}
# ── Key parameters (adjust for speed vs. accuracy) ───────────────────────────
# tau_vec  : quantile levels aggregated by all CQRM variants
# B        : number of null draws for permutation calibration
#            Paper uses B = 10000; set B = 500 here for a fast demo
# M        : jitter replicates per gene per quantile
#            Paper uses M = 50; set M = 10 here for speed
# q_thresh : FDR threshold for FDP/power evaluation
tau_vec  <- c(0.2, 0.4, 0.6, 0.8)
B        <- 500    # increase to 10000 for production
M        <- 10     # increase to 50 for production
q_thresh <- 0.05
```

## Comparison methods

```{r run-comparisons, results='hide'}
cat("DESeq2...\n")
pv_deseq2   <- mydeseq2(count1, count2)

cat("Wilcoxon (DESeq2-normalized)...\n")
pv_wilcoxon <- my.wilcoxon.deseq2(count1, count2)

cat("edgeR...\n")
pv_edger    <- myedger(count1, count2)

cat("limma-voom...\n")
pv_limma    <- tryCatch(my.limmavoom(count1, count2),
                         error = function(e) rep(NA, G))
```

## CQRM methods

All four CQRM functions return **calibrated raw p-values**. Apply
`p.adjust(..., method = "BH")` to obtain FDR-controlled adjusted p-values.

### CQRM.approx — Fast approximation (recommended starting point)

```{r run-approx}
# No jittering: single rq.wfit call per quantile (fastest CQRM variant).
# solve(subsigma) precomputed outside the B-loop.
pv_approx <- my.cqrm.approx(
  count1,
  count2,
  tau_vec = tau_vec,   # quantile levels: 0.2, 0.4, 0.6, 0.8
  B       = B,         # null draws for calibration
  zeta    = 1e-5       # floor for log transformation (numerical stability)
)
adj_approx <- p.adjust(pv_approx, method = "BH")
```

### CQRM.best — Single-quantile MinP

```{r run-best}
# Tests each tau in tau_vec independently; MinP across quantiles.
# Null distribution from independent normals (scalar variance, no matrix solve).
pv_best <- my.cqrm.best(
  count1,
  count2,
  tau_vec = tau_vec,   # quantile levels tested individually
  B       = B,         # null draws for calibration
  M       = M          # jitter replicates per gene per quantile
)
adj_best <- p.adjust(pv_best, method = "BH")
```

### CQRM.region — Contiguous-window MinP

```{r run-region}
# Tests all contiguous windows of k consecutive quantiles; MinP across windows.
# solve(subsigma) precomputed once per window outside the B-loop.
pv_region <- my.cqrm.region(
  count1,
  count2,
  tau_vec = tau_vec,   # base quantile levels
  k       = 2,         # window width (number of consecutive quantiles per test)
  B       = B,         # null draws for calibration
  M       = M          # jitter replicates per gene per quantile
)
adj_region <- p.adjust(pv_region, method = "BH")
```

### CQRM — Full composite test (all 2^K-1 subsets)

```{r run-cqrm}
# Tests all 2^K-1 non-empty subsets of tau_vec; MinP across subsets.
# Null calibrated from MVN null. solve(subsigma) precomputed outside B-loop.
pv_cqrm <- my.cqrm(
  count1,
  count2,
  tau_vec = tau_vec,   # quantile levels (K = 4 => 15 non-empty subsets tested)
  B       = B,         # null draws for calibration
  M       = M,         # jitter replicates per gene per quantile
  seed    = 12345      # RNG seed for reproducibility
)
adj_cqrm <- p.adjust(pv_cqrm, method = "BH")
```

---

# Results

## Adjusted p-value table (top 20 genes by CQRM)

```{r pval-table}
adj_pval_df <- data.frame(
  Gene         = seq_len(G),
  True_DE      = is_true_DE,
  DESeq2       = round(pv_deseq2,   4),
  Wilcoxon     = round(pv_wilcoxon, 4),
  edgeR        = round(pv_edger,    4),
  limma_voom   = round(pv_limma,    4),
  CQRM_Approx = round(adj_approx,  4),
  CQRM_Best   = round(adj_best,    4),
  CQRM_Region = round(adj_region,  4),
  CQRM        = round(adj_cqrm,    4)
)

# Sort by CQRM adj p-value and display top 20
top20 <- head(adj_pval_df[order(adj_pval_df$CQRM), ], 20)

kable(
  top20,
  caption = "Top 20 genes ranked by CQRM adjusted p-value",
  align   = "r",
  row.names = FALSE
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  column_spec(2,  bold = TRUE,  color = ifelse(top20$True_DE, "green", "red")) |>
  column_spec(10, bold = TRUE)
```

*The `True_DE` column is coloured **green** for true positives and **red** for
false positives.*

## FDP / Power summary

```{r fdp-power-table}
compute_fdp_power <- function(adj_pv, true_de, q = 0.05) {
  adj_pv[is.na(adj_pv)] <- 1
  disc  <- which(adj_pv <= q)
  if (length(disc) == 0) return(c(FDP = NA_real_, Power = 0))
  c(
    FDP   = round(sum(!disc %in% which(true_de)) / length(disc), 4),
    Power = round(sum(disc %in%  which(true_de)) / sum(true_de), 4)
  )
}

method_names <- c("DESeq2", "Wilcoxon", "edgeR", "limma-voom",
                  "CQRM-Approx", "CQRM-Best", "CQRM-Region", "CQRM")
adj_list <- list(pv_deseq2, pv_wilcoxon, pv_edger, pv_limma,
                 adj_approx, adj_best, adj_region, adj_cqrm)

fdp_power_df <- do.call(rbind, lapply(seq_along(method_names), function(i) {
  res <- compute_fdp_power(adj_list[[i]], is_true_DE, q = q_thresh)
  data.frame(Method = method_names[i], FDP = res["FDP"], Power = res["Power"])
}))
rownames(fdp_power_df) <- NULL

kable(
  fdp_power_df,
  caption = sprintf("FDP and Power at FDR threshold q = %.2f", q_thresh),
  align   = "r",
  row.names = FALSE
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) |>
  row_spec(which(fdp_power_df$Method %in% c("CQRM","CQRM-Approx","CQRM-Best","CQRM-Region")),
           bold = TRUE, background = "#f0f7ff")
```

## Save results to CSV

```{r save-results}
dir.create("results", showWarnings = FALSE)
write.csv(adj_pval_df,  "results/adj_pval_table.csv",  row.names = FALSE)
write.csv(fdp_power_df, "results/fdp_power_table.csv", row.names = FALSE)
cat("Saved adj_pval_table.csv and fdp_power_table.csv to results/\n")
```

---

# Notes on CQRM Implementation

## Why `solve(subsigma)` is moved outside the B-loop

In all CQRM variants, the null p-value calibration loop draws `B` score
vectors from a multivariate normal distribution with covariance `sigma`.
For each draw `b` and each weight pattern `l`, a chi-squared statistic is
computed as:

```
chi2 = t(subscore) %*% solve(subsigma) %*% subscore
```

Because `sigma` is **fixed** across all `B` draws (it is the analytic
covariance matrix from the last gene), `subsigma_inv <- solve(subsigma[wt,wt])`
can be precomputed **once per weight pattern** before the B-loop:

```r
# Outside the B-loop — computed once:
subsigma_inv <- lapply(seq_len(L), function(l) {
  wt <- wt_logical[[l]]
  solve(sigma[wt, wt, drop = FALSE])
})

# Inside the B-loop — reuse the precomputed inverse:
for (b in seq_len(B)) {
  bscore <- score_null[b, ]
  for (l in seq_len(L)) {
    wt      <- wt_logical[[l]]
    subscore <- bscore[wt]
    subchi2  <- drop(t(subscore) %*% subsigma_inv[[l]] %*% subscore)
    ...
  }
}
```

This reduces `B × L` matrix inversions to `L` inversions, providing a
significant speedup especially for large `B`.

## CQRM variant comparison

| Variant | Subsets tested | Null sigma | Speed |
|---------|---------------|------------|-------|
| `CQRM` | All 2^K-1 non-empty | MVN (matrix) | Slowest |
| `CQRM.region` | Contiguous k-windows | MVN (matrix) | Medium |
| `CQRM.best` | Individual quantiles | Independent normals | Fast |
| `CQRM.approx` | All 2^K-1 non-empty | MVN (matrix) | Fastest† |

†`CQRM.approx` replaces M jitter replicates with a single `rq.wfit` call per
quantile.

---

# Session Info

```{r session-info}
sessionInfo()
```
